Temperature and moisture effects on greenhouse gas emissions from deep active-layer

### 2016-04-15 manuscript/cpcrw_manuscript.Rmd
Incubation ms - discussion additions, some results

Abstract.

@@

Temperature and moisture typically have strong and often interactive influences on soil GHG emissions. A long literature of laboratory incubations, field observations, and meta-analyses has documented increased CO2 fluxes with rising temperature {Davidson, 2006 #1669}{Hashimoto, 2015 #4231}{Treat, 2015 #4217}. GHG responses to wetting and thawing dynamics are much less certain, with substantial variability between studies {Kim, 2012 #2863}. The anaerobic conditions common following permafrost thaw are expected to lower CO~2~ emissions but increase those of CH~4~ {Treat, 2015 #4217;Treat, 2014 #4213}, and interact with soil temperature effects in long-term incubations {Elberling, 2013 #3720}. Decadal warming, and perhaps drying, trends in Alaska {Bieniek, 2014 #4389} will work to counteract these effects, with uncertain consequences for the carbon cycle and climate system.

The goal of this study was to examine how temperature and moisture control C loss and GHG (CO~2~ and CH~4~) emissions from soils sampled from the bottom of the annual active layer--i.e., directly above permafrost--in an Alaskan boreal forest. Most previous studies have focused on surface soils or permafrost soils, neglecting deep active-layer soils that experience fluctuating temperature and, especially, moisture conditions. Such deep but unfrozen soils were identified as subject to strong effects from a two-decade warming experiment in the Alaskan Arctic {Sistla, 2013 #3666}. We also aimed to characterize the chemical and biological properties of these soils as they were subjected to a 100-day incubation at difference temperature and moisture levels. We hypothesized that (i) CO~2~ would be the dominant pathway for C loss in these largely aerobic soils; (ii) soils maintained at field moisture and high temperature would lose more C-CO~2~ than cores incubated at 4˚C, due to increased aerobic and anaerobic microbial activity; and (iii) core CH~4~ fluxes would be sensitive only to temperature, as no anaerobic conditions were planned for the cores.

@@

On 18 August 2015 cores were placed into one of two growth chambers (Conviron CMP6050 Control Systems, Winnipeg, Canada) maintained at 4 and 20 °C temperatures and 70% relative humidity and allowed to equilibrate for two weeks. Starting on 31 August 2015 we measured the cores' mass and GHG (CH~4~ and CO~2~) evolution four times in the first week, then twice per week for the first month, and then once per week for the rest of the 100-day incubation. Throughout the incubation, cores were mounted on 0.05-MPa ceramic plates (Soil Moisture Equipment Corp., Santa Barbara, CA, USA) so that, when the plates were placed in contact with water, water would move up into the cores via capillary action. This was done to moderate the drying of the 20 °C "controlled drought" group. The 4 and 20 °C "field moisture" groups received additional wetting from the top of the cores to maintain their water status.

@@

All data analysis and statistics were performed using R version `r R.version.string` {R Development Core Team, 2016 #4314}. This experiment was run as an "open experiment" {Bond-Lamberty, 2016 #4388} with all analysis code, data (from raw instrument data to final summaries), diagnostics, etc., available at https://github.com/bpbond/cpcrw_incubation. The summarized flux data backing the main results have been archived at [DOI TO BE FILLED IN UPON ACCEPTANCE].

@@

After removal of known problematic data, and exclusion of statistical outliers, CO~2~ flux rates measured during the incubation ranged from `r pn(fluxsummary["flux_µgC_gC_day_min", "CO2"], 2)` µg C g C^-1^ day^-1^ (`r pn(fluxsummary["flux_µgC_g_day_min", "CO2"], 2)` µg C g soil^-1^ day^-1^) to a maximum of `r pn(fluxsummary["flux_µgC_gC_day_max", "CO2"], 2)` (`r pn(fluxsummary["flux_µgC_g_day_max", "CO2"], 2)`), with a mean of `r pn(fluxsummary["flux_µgC_gC_day_mean", "CO2"], 3)` (`r pn(fluxsummary["flux_µgC_g_day_mean", "CO2"], 2)`). CH~4~ rates ranged from `r pn(fluxsummary["flux_µgC_gC_day_min", "CH4"]*1000, 2)` ng C g C^-1^ day^-1^ (`r pn(fluxsummary["flux_µgC_g_day_min", "CH4"]*1000, 2)` ng C g soil^-1^ day^-1^) to a maximum of `r pn(fluxsummary["flux_µgC_gC_day_max", "CH4"]*1000, 2)` (`r pn(fluxsummary["flux_µgC_g_day_max", "CH4"]*1000, 3)`), with a mean of `r pn(fluxsummary["flux_µgC_gC_day_mean", "CH4"]*1000, 2)` (`r pn(fluxsummary["flux_µgC_g_day_mean", "CH4"]*1000, 2)`).

These means conceal considerable variablility over the course of the incubation (**Figures 2 and 3**). In the linear mixed-effects model, CO~2~ was strongly influenced by incubation chamber temperature, core gravimetric water content, and percent soil N (all P < 0.05, and the latter two P < 0.001; **Table 1**). (Note that percent C and percent N were highly correlated (r ≥ 0.98) for these cores. Because percent N was a slightlyl stronger predictor, it was retained in the model while percent C was excluded.) The interaction between water content and soil N was also highly significant (`r pclean(summary(step_co2_lme)$tTable["WC_gravimetric:N_percent", "p-value"])`), with high-N cores having little relationship between water content and CO~2~ flux (data not shown). Methane fluxes were most strongly correlated with percent N, while water content exhibited significant interactions with percent N and DOC (**Table 2**). Neither temperature nor water content was a significant first-order predictor of CH~4~ fluxes, however.

@@

Rises in boreal soil temperatures - who has quantified this? Changes in soil WC much less certain. Thermokarst etc., deepening of active layer - that's the larger implications of this work.

Modeling results highly dependent on water content of soils after permafrost degradation.

Eastern Alaska high latitude warming and forest stress {Ju, 2016 #4320}.

@@

Several studies have measured microbial respiration and GHG fluxes from soils very close to our study site. Morishita et al. {, 2014 #4316} measured gas fluxes in the field at CPCRW and nearby forests, and found CO~2~ production to be correlated with both temperature and moisture, consistent with our results They found however that CH~4~ uptake (no emissions were observed) was driven by temperature only. Waldrop et al. {, 2010 #4380} incubated active-layer and permafrost soils from near Fairbanks, AK, under varied temperature and aerobic conditions, observing Q~10~ values of 9.0 (active layer) and 2.3 (permafrost) from -5 to 5 °C. Flux rates were 0.001-0.10 µmol CH~4~ day^-1^ g^-1^ (~`r pn(0.001 * 16 / 0.012 / 1000, 3)`-`r pn(0.1 * 16 / 0.012 / 1000, 3)` ng C g C^-1^ day^-1^), differing by orders of magnitude between sites but roughly similar to our observed emissions, and ~1-5 µg C-CO2 hr^-1^ g^-1^ (~`r pn(1 * 24 / 0.012, 0)`-`r pn(5 * 24 / 0.012, 0)` µg C g C^-1^ day^-1^), considerably higher than the CO~2~ rates observed from our cores.

More generally,

Treat et al. 2014 also showed that CH4 production from organic permafrost soils has a positive relationship with temperature (cites several studies: Lupasco et al. 2012, etc.)
*	Treat et al. 2015. Meta-analysis of anaerobic incubations.  Mineral soils 20-100 cm depth had a mean CO2 production (ug CO2-C /g C/day) of 100.6 +/- 10.3 (33.2 +/- 3.7 per g DW)  (7840 per m2/d). All soils < 10C were 67 +/- 16 and > 20C was 172 +/- 17. Dry soils 68 +/- 8 and Fluctuating = 364 +/- 60 and inundated = 105 +/- 11.

*	Xue et al. 2016: active layer, contained organic horizon: soil moisture increased in warmed plots: ~50% moisture compared to ~45% (p < 0.05. C Flux (g/m2) from ecosystem respiration was greater in warmed plots (~500 g/m2) compared to control plots (~400) (p < 0.05). Authors remarked that soil moisture/climate was important factor in shifting functional composition of microbial community (GeoChip), esp. with anaerobic activity.

*	Schaufler et al. 2010: 33.7 +/- 2.7C02-C (mg C/ m2/ hr) averaged across all temperature (5-20C) and moisture (20-80% WFPS) in forested, boreal site, with 33.7 +/- 2.7.  CH4-C = -30.2 +/- 4.5. Optimum moisture content was 40% WFPS for this site – or highest CO2 evolved at this moisture content. Carbon dioxide emissions were linearly correlated with temperature (p = 0.000) and exponentially correlated with WFPS (p = 0.000). CH4 uptake (oxidation) was highest in the northern sites (included boreal site), and was also the driest soil moisture conditions. -44.8 +/- 5.5. ug CH4-C/m2/hr and decreased with increasing soil moisture.  Effect of WFPS on CH4 oxidation, exponential, p  = 0.000.

*	Wu et al. 2010. Mongolian steppe 900-1400msl. Soils = chernozems, but there was a gleyosol. Soil moisture and temperature influenced CH4. Soil drought limited CH4 uptake at all 15C. CO2 in meadow sites – increased with temperature (< 30 mg C/m2/hr @ 5C to 90 at 15C) and with moisture (<30-50 in dry treatment from 5-15C, and 50-90 in high rain treatment at 5-15C).

Seeta’s paper: Mineral soil: greenhouse treatment (warming treatment) = 0.646 ± 0.04 GREATER THAN control = 0.43 ± 0.05 C-CO2 per g soil per day in 14 day incubation ( α ≤ 0.1). Mineral soil also had greater active biomass and extractable-OM). (Sistla et al. 2013 Nature).

**Temperature versus moisture sensitivity, and CO~2~ versus CH~4~**

Why would we expect to see greater CO2 evolved in higher temps?

*	Increased temp (to a certain degree) increases enzyme kinetics (i.e. Arrhenius equation) (Kirschbaun  1995, 2006 for SOM-temp, Wallenstein/Allison/Schimel for enzyme-temp, etc. so many others)
*	Increased temperature enhances the metabolic rate of microorganisms – thus increasing the production of extracellular enzymes (state somewhere prior that extracellular enzymes catalyze fundamental biogeochemical reactions in SOC decomposition and nutrient cycling). Cusack et al. (2010) suggests temperature-sensitivity of decomposition may be indirectly related to temperature dependence of enzyme production AND turnover. 
*	Carbon use efficiency (CUE), a measure of how well a microorganism can assimilate C into biomass relative to how much C is lost as CO2,  has been shown to vary with temperature (CITE – Lopez-Moran 2007; Price and Sowers 2004 !) with decreases in CUE at higher temperatures leading to greater CO2 production during SOM decomposition (Steinweg et al. 2008). Cite Balser and Wixon 2009 in regards to microbial-mediated effects on temp-sensitivity of C decomposition. 
*	Higher temperatures enhance desorption rates of OM with minerals (ten Hulscher and Cornelissen 1996; Kalbitz et al. 2000 – others) making more OC available in solution (Xu & Saiers, 2010; Plante et al. 2009), and thus accessible for adjacent, competent microorganisms

Note high CO2:CH4 ratios, compare with Treat et al. and other studies.

Increased temperature alone may not drive C mineralization rates in forest-mineral soils (Giardina and Ryan, 2000), but its corresponding increase in soil moisture content can lead to changes in microbial community structure and enhance anaerobic activity that may amplify CO2, CH4 and N2O fluxes (Xue et al, 2016).

@@

Incubation experiments can only go so far though...
Such laboratory studies have weaknesses {Podrebarac, 2016 #4365}, but their controlled environments provide an important way to elucidate the key mechanisms controlling GHG from high-latitude soils {Schuur, 2015 #4319}.

@@

Check out:
Soil microbial respiration rate and temperature sensitivity along a north-south forest transect in eastern China: Patterns and influencing factors
http://onlinelibrary.wiley.com/doi/10.1002/2015JG003217/abstract

TODO: look at Podrebarac (2016) - Soils isolated during incubation underestimate temperature sensitivity of respiration and its response to climate history

@@

APS comment: Did the controlled drought treatment match the 4C drought treatment in terms of rate of water loss (for both gravimetric AND volumetric)? Since the controlled drought treatment was meant to control for increased water loss with the higher temp, I want to make sure that it really was doing that. From this graph, it “looks” like it followed the same trend as the 20C drought treatment.  TODO

@@

APS comment: Was difference in cum CO2 between field moist cores and drought cores the same for both temperatures??? Granted cum Co2 was greater overall in 20C, but interesting that the difference between treatments appear equal in both temperatures. This is a bit difference from Treat et al. 2014 (granted they had permafrost/peat soils) – but Treat reported larger differences between wet/dry (aerobic/anaerobic) treatments at 20C but no differences at -0.5C, but explain this due to frozen water limiting microbial respiration.      TODO

>>>>

High latitude ecosystems are being subjected to rapid climatic changes, with rising air temperatures and fires expected to drive permafrost degradation and soil carbon (C) cycle changes. How these soils' greenhouse gas emissions will respond to changes in temperature, moisture, and vegetation is highly uncertain. We used a laboratory experiment to examine how temperature and moisture control CO~2~ and CH~4~ emissions from soils sampled from the bottom of the annual active layer in an Alaskan boreal forest. Gas emissions were measured from thirty cores, subjected to two temperatures and either field moisture conditions or drought, over a 100-day incubation. TODO...

@@

Temperature and moisture typically have strong and often interactive influences on soil GHG emissions. A long literature of laboratory incubations, field observations, and meta-analyses has documented increased CO2 fluxes with rising temperature {Davidson, 2006 #1669}{Hashimoto, 2015 #4231}{Treat, 2015 #4217}. GHG responses to wetting and thawing dynamics are much less certain, with substantial variability between studies {Kim, 2012 #2863}. The anaerobic conditions common following permafrost thaw are expected to lower CO~2~ emissions but increase those of CH~4~ {Treat, 2015 #4217;Treat, 2014 #4213}, and interact with soil temperature effects in long-term incubations {Elberling, 2013 #3720}. More uniformly across the landscape, however, decadal warming, and perhaps drying, trends in Alaska {Bieniek, 2014 #4389} will work to counteract these effects, with uncertain consequences for the carbon cycle and climate system.

The goal of this study was to examine how temperature and moisture control C loss and GHG (CO~2~ and CH~4~) emissions from soils sampled from the bottom of the annual active layer--i.e., directly above permafrost--in an Alaskan boreal forest. Most previous studies have focused on surface soils or permafrost soils, neglecting deep active-layer soils that experience fluctuating temperature and, especially, moisture conditions. Such deep but unfrozen soils were identified as subject to strong effects from a two-decade warming experiment in the Alaskan Arctic {Sistla, 2013 #3666}. We also aimed to characterize the chemical and biological properties of these soils as they were subjected to a 100-day incubation at different temperatures, subjecting some cores to drying treatments. We hypothesized that (i) CO~2~ would be the dominant pathway for C loss in these largely aerobic soils; (ii) soils maintained at field moisture and high temperature would lose more C-CO~2~ than cores incubated at 4˚C, due to increased aerobic and anaerobic microbial activity; and (iii) core CH~4~ fluxes would be sensitive only to temperature, as no anaerobic conditions were planned for the cores.

@@

On 18 August 2015 cores were placed into one of two growth chambers (Conviron CMP6050 Control Systems, Winnipeg, Canada) maintained at 4 and 20 °C temperatures and 70% relative humidity and allowed to equilibrate for two weeks. Starting on 31 August 2015 we measured the cores' mass and GHG (CH~4~ and CO~2~) evolution four times in the first week, then twice per week for the first month, and then once per week for the rest of the 100-day incubation. Throughout the incubation, cores were mounted on porous ceramic plates (Soil Moisture Equipment Corp., Santa Barbara, CA, USA) so that, when the plates were placed in contact with water, water would move up into the cores via capillary action. The "drought" cores were mounted on dry plates, but not allowed to drop below 5% water content. When necessary, cores received additional wetting from the top to maintain their water status at the desired level.

@@

All data analysis and statistics were performed using R version `r R.version.string` {R Development Core Team, 2016 #4314}. This experiment was run as an "open experiment" {Bond-Lamberty, 2016 #4388} with all analysis code, data (from raw instrument data to final summaries), diagnostics, etc., available at https://github.com/bpbond/cpcrw_incubation. The summarized flux data backing the main results have been archived at [DOI to be filled in on acceptance].

@@

After removal of known problematic data, and exclusion of statistical outliers, CO~2~ flux rates measured during the incubation ranged from `r pn(fluxsummary["flux_µgC_gC_day_min", "CO2"], 1)` µg C g C^-1^ day^-1^ (`r pn(fluxsummary["flux_µgC_g_day_min", "CO2"], 1)` µg C g soil^-1^ day^-1^) to a maximum of `r pn(fluxsummary["flux_µgC_gC_day_max", "CO2"], 1)` (`r pn(fluxsummary["flux_µgC_g_day_max", "CO2"], 2)`), with a mean of `r pn(fluxsummary["flux_µgC_gC_day_mean", "CO2"], 1)` (`r pn(fluxsummary["flux_µgC_g_day_mean", "CO2"], 1)`). CH~4~ rates ranged from `r pn(fluxsummary["flux_µgC_gC_day_min", "CH4"]*1000, 2)` ng C g C^-1^ day^-1^ (`r pn(fluxsummary["flux_µgC_g_day_min", "CH4"]*1000, 2)` ng C g soil^-1^ day^-1^) to a maximum of `r pn(fluxsummary["flux_µgC_gC_day_max", "CH4"]*1000, 2)` (`r pn(fluxsummary["flux_µgC_g_day_max", "CH4"]*1000, 3)`), with a mean of `r pn(fluxsummary["flux_µgC_gC_day_mean", "CH4"]*1000, 2)` (`r pn(fluxsummary["flux_µgC_g_day_mean", "CH4"]*1000, 2)`).

These means conceal considerable variablility over the course of the incubation (**Figures 2 and 3**). In the linear mixed-effects model, CO~2~ was strongly influenced by incubation chamber temperature, core gravimetric water content, and percent soil N (all P < 0.05, and the latter two P < 0.001; **Table 1**). (Note that percent C and percent N were highly correlated (r ≥ 0.98) for these cores. Because percent N was a slightlyl stronger predictor, it was retained in the model while percent C was excluded.) The interaction between water content and soil N was also highly significant (`r pclean(summary(step_co2_lme)$tTable["WC_gravimetric:N_percent", "p-value"])`), with high-N cores having little relationship between water content and CO~2~ flux (data not shown). "Drought" cores' cumulative production was `r pn(drought_effect["4", "Reduction"], 0)`% (4 °C) and `r pn(drought_effect["20", "Reduction"], 0)`% (20 °C) lower than the cores kept at field moisture.

Methane fluxes were most strongly correlated with percent N, while water content exhibited significant interactions with percent N and DOC (**Table 2**). Neither temperature nor water content was a significant first-order predictor of CH~4~ fluxes, however.

@@

Rises in boreal air temperatures, and unpredictable precipitation changes, will warm and dry many soils, increase vegetation stress {Ju, 2016 #4320}, degrade permafrost and deepen the active layer {Schuur, 2015 #4319}. The resulting changes in the soil carbon cycle may already be visible {Bond-Lamberty, 2012 #3480} but are highly uncertain going forward, increasing the need for controlled laboratory studies of GHG emission dynamics from high-latitude soils.

@@

Several studies have measured microbial respiration and GHG fluxes from soils very close to our study site. Morishita et al. {, 2014 #4316} measured gas fluxes in the field at CPCRW and nearby forests, and found CO~2~ production to be correlated with both temperature and moisture, consistent with our results They found however that CH~4~ uptake (no emissions were observed) was driven by temperature only. Waldrop et al. {, 2010 #4380} incubated active-layer and permafrost soils from near Fairbanks, AK, under varied temperature and aerobic conditions, observing Q~10~ values of 9.0 (active layer) and 2.3 (permafrost) from -5 to 5 °C; these values are higher than we observed, consistent with the lower temperature range {Hamdi, 2013 #3664}. Waldrop et al. {, 2010 #4380} observed flux rates of 0.001-0.10 µmol CH~4~ day^-1^ g^-1^ (~`r pn(0.001 * 16 / 0.012 / 1000, 3)`-`r pn(0.1 * 16 / 0.012 / 1000, 3)` ng C g C^-1^ day^-1^), differing by orders of magnitude between sites (but roughly similar to our observed CH~4~ emissions), and ~1-5 µg C-CO2 hr^-1^ g^-1^ (~`r pn(1 * 24 / 0.012, 0)`-`r pn(5 * 24 / 0.012, 0)` µg C g C^-1^ day^-1^), considerably higher than the CO~2~ rates observed from our cores. In an incubation of active-layer Alaskan permafrost peats, Treat et al. {, 2014 #4213} found CO~2~ and CH~4~ emissions to be strongly correlated with temperature and moisture. Finally, during the first 100 days of a year-long incubation of Fairbanks-area 0-10 cm mineral soils, Neff and Hooper {, 2002 #4390} observed fluxes of ~`r pn(0.2 * 1000 * 12/44, 0)`-`r pn(1.5 * 1000 * 12/44, 0)` µg C-CO~2~ g C^-1^ day^-1^, in line with the results here.

More generally, in a pan-Arctic synthesis of anaerobic soil incubations, Treat et al. {, 2015 #4217} reported mean CO~2~ rates of 47 (all mineral soils) and 101 (for 20-100 cm soils) µg C-CO~2~ g C^-1^ day^-1^, somewhat lower than our aerobic incubation results. Weiss et al. {, 2015 #4317} found CO~2~ production from Yedoma permafrost samples to be correlated with both percent C and N, consistent with our results (**Table 1**). TODO.

**Temperature versus moisture sensitivity**

Warming usually increases soil GHG fluxes, for example at depth in a long-term Arctic tundra experiment {Sistla, 2013 #3666}, as increased temperatures increment enzyme kinetics, enhance the production of extracellular enzymes, and enhance desorptions rates of organic matter from minerals. A key question, for both experimentalists and modelers {Falloon, 2011 #3045}, is to what degree such soils' emissions could by constrained by their moisture status--in particular, limited by low soil moisture driven by increases in high-latitude temperatures, vapor pressure deficit, and potentially precipitation changes.

Our results suggest that moisture limitation could exert a large effect on CO~2~ production for deep active-layer soils (**Figure 4**): "drought" cores' cumulative production was `r pn(drought_effect["4", "Reduction"], 0)`% (4 °C) and `r pn(drought_effect["20", "Reduction"], 0)`% (20 °C) lower than the cores kept at field moisture. This effect was highly significant, and suggests that moisture limitations could exert a significant constraint on deep active-layer soils as they slowly warm. Such moisture constraints are thought to be already exerting effects on vegetation and soil fluxes at large scales {Ju, 2016 #4320}{Bond-Lamberty, 2012 #3480}, but our understanding of the interactive effects involved is poor.

In contrast, the temperature sensitivities observed in this experiment are low (all less than 2.0, even when controlling for changes in soil moisture). Such values are not unprecedented in comparison to a wide range of other laboratory soil incubations {Hamdi, 2013 #3664}. It is important to note, however, that even if increased temperature alone may not always drive C mineralization rates in forest-mineral soils {Giardina, 2000 #2159}, but it is frequently linked with increases in soil moisture content can lead to changes in microbial community structure, amplifying GHG fluxes (Xue et al, 2016).  TODO: PEYTON - WHAT IS THIS CITATION?

**Methane**

We observed very low, but consistently positive, CH~4~ production from these unsaturated mineral soils. This is contrast to many field studies, e.g. Schaufler et al. {, 2010 #4391} who found consistent CH~4~ uptake (oxidation) in dry conditions at a boreal site.

Treat et al. found a median CO~2~:CH~4~ production ratios of 387 for boreal sites, far lower than (but consistent with) our observed ratios of several million.

@@

Incubation experiments can only go so far though...

Laboratory soil incubations have weaknesses, for example in what can be inferred about temperature sensitivity {Podrebarac, 2016 #4365}{Hamdi, 2013 #3664}, but their controlled environments provide an important way to elucidate the key mechanisms controlling GHG from high-latitude soils {Schuur, 2015 #4319}.

@@

Necessity of integrated lab-field-modeling experiments...

[//]: (comment comment comment)

@@

@@
